data_collation

### 2024-05-23 scripts/data_collation.Rmd
redid swath, cleaned quaad, and species meta

# output updated dataframe
write.csv(kelp_data, "output/cleaned_data/kelp.data.clean.csv", row.names = FALSE)
```

@@

#combine kelp and morph daat
morph_data <- left_join(morph_data, kelp_data)

@@

@@

# Fish data is cleaned and ready

@@

# combine updated dataframes
fish_data <- left_join(fish, fish_data)

@@

# look at fish_Aggregated later - Austin, this is kelp density possibly?

@@

# make columns lowercase
# Get the current column names
current_names <- colnames(UPC_data)

# Convert the column names to lowercase while preserving the first letter
new_names <- sapply(current_names[1:13], function(x) {
  paste0(toupper(substring(x, 1, 1)), tolower(substring(x, 2)))
})

# Assign the new column names to the first six columns of the dataframe
colnames(UPC_data)[1:13] <- new_names

@@

# swath dataframe has species names as columns in wide format, need to change names to species codes to combine with other dataframes

# Load species names and species code data frame
species.metadata <- read.csv("data/species_code_metadata.csv")

# Function to rename columns using metadata
# Get the unique column names from the swath dataframe
column_names <- colnames(swath)

# Create a lookup table with name-to-code mappings
name_to_code <- setNames(species.metadata$code, species.metadata$name)

# Rename the columns using the lookup table
  new_column_names <- sapply(column_names, function(name) {
    if (name %in% names(name_to_code)) {
      return(name_to_code[name])
    } else {
      return(name)
    }
  })

# Rename the columns
colnames(swath) <- new_column_names

# Convert data in numeric columns to numeric
swath.con <- swath %>%
  mutate(across(AGH:BLOA, as.numeric))

# Melt only columns AGH:WHEL into long format
swath_long <- swath.con %>%
  pivot_longer(cols = AGH:BLOA, names_to = "SP_CODE", values_to = "value")

# Need to now move values to binned swatch categories as columns
# Pivot the dataframe into wide format
swath_binned <- swath_long %>%
  pivot_wider(names_from = swath, values_from = value)

# make columns lowercase
# Get the current column names
current_names <- colnames(swath_binned)

# Convert the column names to lowercase while preserving the first letter
new_names <- sapply(current_names[1:6], function(x) {
  paste0(toupper(substring(x, 1, 1)), tolower(substring(x, 2)))
})

# Assign the new column names to the first six columns of the dataframe
colnames(swath_binned)[1:6] <- new_names

@@

swath_count <- swath_count %>%
  rename(
    "0-20 IN" = "X0.20.IN",
    "20-40 IN" = "X20.40.IN",
    "40-20 OFF" = "X40.20.OFF",
    "20-0 OFF" = "X20.0.OFF"
  )

FtWeatherill_swath_2016 <- FtWeatherill_swath_2016 %>%
  rename(
    "0-20 IN" = "X0.20.IN",
    "20-40 IN" = "X20.40.IN",
    "40-20 OFF" = "X40.20.OFF",
    "20-0 OFF" = "X20.0.OFF")

# Remove four columns in a specific order swath_2018
swath_2018 <- swath_2018 %>%
  select(-12:-15)

# Remove four columns in a specific order swath_count
swath_count <- swath_count %>%
  select(-12:-15)

# make columns lowercase
# Get the current column names
#current_names <- colnames(FtWeatherill_swath_2016)

# Convert the column names to lowercase while preserving the first letter
#new_names <- sapply(current_names[1:6], function(x) {
#  paste0(toupper(substring(x, 1, 1)), tolower(substring(x, 2)))
#})

# Assign the new column names to the first six columns of the dataframe
#colnames(FtWeatherill_swath_2016)[1:6] <- new_names

# use rbind.fill() method in R to combine data frames with different columns
swath_data <- rbind.fill(swath_binned, swath_2018, swath_count, FtWeatherill_swath_2016)

# Rename ft. wetherill to fort wetherill
swath_data$Site[swath_data$Site == "Ft.Weatherill"] <- "Fort Wetherill"

@@

# To do: species codes transform

# use rbind.fill() method in R to combine data frames with different columns
QUAD_data <- rbind.fill(QUAD, QUAD_2018, FtWeatherill_QUAD_2017, FtWeatherill_QUAD_2016_TALBENHORIN)

@@

>>>>

@@

#combine kelp and morph daat
morph_data <- rbind.fill(morph_data, kelp_data)

@@

# Kelp COVER data 
```{r}
# kelp cover dataframe has species names as columns in wide format, need to change names to species codes to combine with other dataframes
# Load species names and species code data frame
species.metadata <- read.csv("data/species_code_metadata.csv")

# Remove the first column from Kelp_COVER if it's unnecessary
Kelp_COVER <- Kelp_COVER[ , -1 ]

# Function to rename columns using metadata
# Get the unique column names from the Kelp_COVER dataframe
column_names.kelp <- colnames(Kelp_COVER)

# Create a lookup table with name-to-code mappings
name_to_code.kelp <- setNames(species.metadata$code, species.metadata$name)

# Rename the columns using the lookup table
new_column_names.kelp <- sapply(column_names.kelp, function(name) {
  if (name %in% names(name_to_code.kelp)) {
    return(name_to_code.kelp[name])
  } else {
    return(name)
  }
})

# Rename the columns in Kelp_COVER
colnames(Kelp_COVER) <- new_column_names.kelp

# Melt only columns AGH:WHEL into long format
Kelp_COVER_long <- Kelp_COVER %>%
  pivot_longer(cols = AGH:CLT, names_to = "SP_CODE", values_to = "value")

# output updated dataframe
write.csv(Kelp_COVER_long, "output/cleaned_data/kelp.cover.data.clean.csv", row.names = FALSE)

@@

@@

# combine updated dataframes
fish_data <- rbind.fill(fish, fish_data)

@@

@@

@@

@@

# remove columns with observer and unneeded data
swath_2018 <- swath_2018[ -c(12:15) ]

# Make all columns capitalized
colnames(swath_2018) <- toupper(colnames(swath_2018))

#rename in and out size classes
Swath_count <- Swath_count %>%
  rename(
    "0-20 IN" = "X0.20.IN",
    "20-40 IN" = "X20.40.IN",
    "40-20 OFF" = "X40.20.OFF",
    "20-0 OFF" = "X20.0.OFF"
  )

# remove columns with observer and unneeded data
Swath_count <- Swath_count[ -c(12:15) ]

# Make all columns capitalized
colnames(Swath_count) <- toupper(colnames(Swath_count))

# remove columns with observer and unneeded data
FtWeatherill_Swath_2016 <- FtWeatherill_Swath_2016[ -c(9:14) ]

# Renaming values in the 'SIDE' column
FtWeatherill_Swath_2016 <- FtWeatherill_Swath_2016 %>%
  mutate(SIDE = case_when(
    SIDE == "IN 20" ~ "0-20 IN",
    SIDE == "OFF 40" ~ "40-20 OFF",
    SIDE == "IN 40" ~ "20-40 IN",
    SIDE == "OFF 20" ~ "20-0 OFF",
    TRUE ~ SIDE))  # Keep other values unchanged

# Pivoting the data from long to wide format
FtWeatherill_Swath_2016_wide <- FtWeatherill_Swath_2016 %>%
  pivot_wider(
    names_from = SIDE,
    values_from = COUNT)

# use rbind.fill() method in R to combine data frames with different columns
swath_data <- rbind.fill(swath_2018, Swath_count, FtWeatherill_Swath_2016_wide)

# Rename ft. wetherill to fort wetherill
swath_data$SITE[swath_data$SITE == "Ft.Weatherill"] <- "Fort Wetherill"

# Rename ft. wetherill to fort wetherill
swath_data$SITE[swath_data$SITE == "Ft Wetherill"] <- "Fort Wetherill"

@@

#make QUAD data long
# Function to rename columns using metadata
# Get the unique column names from the QUAD ataframe
column_names.quad <- colnames(QUAD)

# Create a lookup table with name-to-code mappings
name_to_code.quad <- setNames(species.metadata$code, species.metadata$name)

# Rename the columns using the lookup table
new_column_names.quad <- sapply(column_names.quad, function(name) {
  if (name %in% names(name_to_code.quad)) {
    return(name_to_code.quad[name])
  } else {
    return(name)
  }
})

# Rename the columns in Kelp_COVER
colnames(QUAD) <- new_column_names.quad

# Melt only columns AGH:WHEL into long format
QUAD_long <- QUAD %>%
  pivot_longer(cols = AGH:Oar.Weed.Total, names_to = "SP_CODE", values_to = "COUNT")

# use rbind.fill() method in R to combine data frames with different columns
QUAD_data <- rbind.fill(QUAD_long, QUAD_2018, FtWeatherill_QUAD_2017, FtWeatherill_QUAD_2016_TALBENHORIN)

#remove 131-138 columns for observers not needed information
QUAD_data <- QUAD_data[ -c(14:17) ]

@@

# Rename Kings Beach to King's Beach
QUAD_data$SITE[QUAD_data$SITE == "Kings Beach"] <- "King's Beach"

# output updated dataframe
write.csv(QUAD_data, "output/cleaned_data/QUAD.data.clean.csv", row.names = FALSE)
